---
title: "Quantitative genetics using the sommer package"
author: "Giovanny Covarrubias-Pazaran"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantitative genetics using the sommer package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The sommer package was developed to provide R users a powerful and reliable multivariate mixed model solver for different genetic and non-genetic analysis in diploid and polyploid organisms. This package allows the user to estimate variance components for a mixed model with the advantage of specifying the variance-covariance structure of the random effects, specify heterogeneous variances, and obtain other parameters such as BLUPs, BLUEs, residuals, fitted values, variances for fixed and random effects, etc. 

The package is focused on problems of the type p > n related to genomic prediction (hybrid prediction & genomic selection) and GWAS analysis, although general mixed models can be fitted as well. The package provides kernels to estimate additive (`A.mat`), dominance (`D.mat`), and epistatic (`E.mat`) relationship matrices that have been shown to increase prediction accuracy under certain scenarios or simply to estimate the variance components of such. The package provides flexibility to fit other genetic models such as full and half diallel models as well.

Vignettes aim to provide several examples in how to use the sommer package under different scenarios in breeding and genetics. We will spend the rest of the space providing examples for: 

1) Heritability ($h^2$) calculation
2) Specifying heterogeneous variances in mixed models
3) Using the pin calculator
4) Half and full diallel designs
5) Genomic selection
6) Single cross prediction
7) Multivariate genetic models and genetic correlations


## Background

The core of the package are the `mmer2` (formula-based) and `mmer` (matrix-based) functions which solve the mixed model equations. The functions are an interface to call the `NR` Direct-Inversion Newton-Raphson (Tunnicliffe 1989; Gilmour et al. 1995; Lee et al. 2015) or the `EMMA` efficient mixed model association algorithm (Kang et al. 2008). 

<br>

The univariate mixed models that sommer can solve have the form:

<br>

$y = X\beta+Zu+\epsilon$   

<br>
or   
<br>

$y = X\beta+Zu_1+...+Zu_i+\epsilon$ 

<br>

where:

<br>

X is an incidence matrix for fixed effects

Z is an incidence matrix for random effects

$\beta$ is the vector for BLUEs of fixed effects

u is the vector for BLUPs of random effects

$\epsilon$ are the residuals

<br>

The variance of the response is known to be the random part of the model:

<br>

Var(y) = Var($Zu+\epsilon$) = ZGZ + R = V

<br>

and with

<br>

u ~ MVN(u, G)

$\epsilon$ ~ MVN(u, R)

<br>

When multiple random effects are present the Z matrix becomes the column binding of each of the $Z_i$ matrices for the i random effects. And the G matrix becomes the diagonal binding of each of the variance covariance structures (K matrices) for the random effects:

<br>

$$\mathbf{Z} = \left[\begin{array}
{rrr}
Z_1 & ... & Z_i \\
\end{array}\right]
$$

$$\mathbf{G} = \left[\begin{array}
{rrr}
K_1 \sigma^2_u1 & 0 & 0 \\
0 & ... & 0 \\
0 & 0 & K_i \sigma^2_ui
\end{array}\right]
$$

Since version 2.0 sommer can handle multivariate models. These have the form:

<br>

$Y = X\beta + Zu + \epsilon$ 

<br>

with:

<br>

$$\mathbf{Y} = \left[\begin{array}
{r}
y_1 \\
y_2 \\
... \\
y_t \\
\end{array}\right]
$$

<br>

$$\mathbf{X} = \left[\begin{array}
{rrr}
X & ... & ... \\
... & ... & ...\\
... & ... & X \\
\end{array}\right]
$$

<br>

$$\mathbf{V} = \left[\begin{array}
{rrr}
Z_1 G_1 Z_1' + ... + Z_1 R_1 Z_1' & ... & Z_1 H_1 Z_t' + ... + Z_1 S_1 Z_t' \\
 ... & ... & ...\\
Z_t H_1 Z_1' + ... + Z_t S_1 Z_1' & ... & Z_t G_1 Z_t' + ... + Z_t R_1 Z_t' \\
\end{array}\right]
$$


<br>

for 't' traits, where G are H are variance and covariance matrices among random effects for the "t" trait, and R and S are variance and covariance matrices among residuals. Here R=S=I$\sigma_\epsilon$ , where I is an identity matrix. We can specify the covariance matrices. BLUPs will also be corrected for such covariances usually leading to more accurate predictions.

<br>

The function `mmer` takes the Zs and Ks for each random effect and construct the neccesary structure inside and estimates the variance components by ML/REML using any of the 4 methods available in sommer. The `mmer2` function is enabled to work in a model-based fashion so user don't have to build the Z's and K matrices. Please refer to the canonical papers listed in the Literature section to check how the algorithms work. We have tested widely the methods to make sure they provide the same solution when the likelihood behaves well but for complex problems they might lead to slightly different answers. If you have any concern please contact me at cova_ruber@live.com.mx.

In the following section we will go in detail over several examples on how to use mixed models in univariate and multivariate case and their use in quantitative genetics.

<br>

## 1) Marker and non-marker based heritability calculation 

The heritability is one of the most popular parameters in the breeding and genetics community. The heritability is usually estimated as narrow sense ($h^2$; only additive variance in the numerator $\sigma^2_A$), and broad sense ($H^2$; all genetic variance in the numerator $\sigma^2_G$).

In a classical experiment with no molecular markers, special designs are performed to estimate and disect the additive ($\sigma^2_A$) and dominance ($\sigma^2_D$) variance along with environmental variability. Designs such as generation analysis, North Carolina designs are used to disect $\sigma^2_A$ and $\sigma^2_D$ to estimate the narrow sense heritability ($h^2$). When no special design is available we can still disect the genetic variance ($\sigma^2_G$) and estimate the broad sense heritability. In this example we will show the broad sense estimation which doesn't use covariance structures for random effects. For big models with no covariance structures, sommer's direct inversion is a bad idea to use but we will show anyways how to do it, for very sparse models we recommend using the lmer function from the lme4 package from Douglas Bates or change to the EM algorithm which uses MME-based algorithms.

The dataset has 41 potato lines evaluated in 5 locations across 3 years in an RCBD design. We show how to fit the model and extract the variance components to calculate the $h^2$.

```{r}
library(sommer)
data(h2)
head(h2)

ans1 <- mmer2(y~1, 
              random = ~Name + Env + Name:Env + Block,
              rcov = ~units,
              data=h2, silent = TRUE)

suma <- summary(ans1)
n.env <- length(levels(h2$Env))
pin(ans1, h2 ~ V1 / ( V1 + (V3/n.env) + (V5/(2*n.env)) ) )

```

The same model can be fitted with the `mmer` function that is actually used by the `mmer2` function in the background. This is just to show that you can create your customized matrices and use the mixed model solver. This is how you would do it:

```{r}
library(sommer)
data(h2)
head(h2)
Z1 <- model.matrix(~Name-1, h2)
Z2 <- model.matrix(~Env-1, h2)
Z3 <- model.matrix(~Env:Name-1, h2)
Z4 <- model.matrix(~Block-1, h2)
ETA <- list(name=list(Z=Z1),env=list(Z=Z2),name.env=list(Z=Z3),block=list(Z=Z4))
y <- h2$y
ans1 <- mmer(Y=y, Z=ETA, silent = TRUE)
vc <- ans1$var.comp
```

Recently with markers becoming cheaper, thousand of markers can be run in the breeding materials. When markers are available, an special design is not neccesary to disect the additive genetic variance. The availability of the additive, dominance and epistatic relationship matrices allow us to estimate $\sigma^2_A$, $\sigma^2_D$ and $\sigma^2_I$.

Assume you have a population, and a similar model like the one displayed previously has been fitted. Now we have BLUPs for the genotypes but in addition we have genetic markers.

```{r}
data(CPdata)
CPpheno$idd <-CPpheno$id; CPpheno$ide <-CPpheno$id
### look at the data
head(CPpheno)
CPgeno[1:5,1:4]
## fit a model including additive and dominance effects
A <- A.mat(CPgeno) # additive relationship matrix
D <- D.mat(CPgeno) # dominance relationship matrix
E <- E.mat(CPgeno) # epistatic relationship matrix

ans.ADE <- mmer2(color~1, 
                 random=~g(id) + g(idd) + g(ide), 
                 rcov=~units,
                 G=list(id=A,idd=D,ide=E), 
                 silent = TRUE, data=CPpheno)
suma <- summary(ans.ADE)$var.comp.table
(H2 <- sum(suma[1:3,1])/sum(suma[,1]))
(h2 <- sum(suma[1,1])/sum(suma[,1]))

```

In the previous example we showed how to estimate the additive ($\sigma^2_A$), dominance ($\sigma^2_D$), and epistatic ($\sigma^2_I$) variance components based on markers and estimate broad ($H^2$) and narrow sense heritability ($h^2$). Notice that we used the `g()` function which indicates that the random effect inside the parenthesis (i.e. id, idd or ide) has a covariance matrix (A, D, or E), that will be specified in the G argument in the form of a list and using the name of the random efect to allow the program to recognize which variance covariance matrix belongs to each random effect. Please DO NOT provide the inverse but the original covariance matrix. This is why we have called the function `g()` and no `giv()` as the popular software asreml.

Just to show one more time that you can use your own matrices we will repeat the same calculation using the `mmer` function:

```{r}
data(CPdata)
### look at the data
head(CPpheno)
CPgeno[1:5,1:4]
## fit a model including additive and dominance effects
Z1 <- model.matrix(~id-1, CPpheno); colnames(Z1) <- gsub("id","",colnames(Z1))
A <- A.mat(CPgeno) # additive relationship matrix
D <- D.mat(CPgeno) # dominance relationship matrix
E <- E.mat(CPgeno) # epistatic relationship matrix
y <- CPpheno$color

ETA <- list(id=list(Z=Z1,K=A),idd=list(Z=Z1,K=D),ide=list(Z=Z1,K=E))
ans.ADE <- mmer(Y=y, Z=ETA, silent = TRUE)
ans.ADE$var.comp
```

## 2) Specifying heterogeneous variances in univariate models

Very often in multi-environment trials, the assumption that genetic variance is the same across locations may be too naive. Because of that, specifying a general genetic component and a location specific genetic variance is the way to go. Although the function 'mmer' implemented in sommer can be used to do that, can be quite cumbersome and messy to create the incidence and variance covariance matrices for fitting those models. For that reason the function 'mmer2' was added to the package to make such models easier to fit.

We estimate variance components for $GCA_2$ and $SCA$ specifying the variance structure.

```{r, fig.show='hold'}
data(cornHybrid)
hybrid2 <- cornHybrid$hybrid # extract cross data
head(hybrid2)
### fit the model
modFD <- mmer2(Yield~1, 
               random=~ at(Location,c("3","4")):GCA2, 
               rcov= ~ at(Location):units,
               data=hybrid2, silent = TRUE)
summary(modFD)
```

In the previous example we showed how the `at` function is used in the `mmer2` solver. By using the `at` function you can specify that i.e. the GCA2 has a different variance in different Locations, in this case locations 3 and 4, but also a main GCA variance. This is considered a CS + DIAG (compound symmetry + diagonal) model.

In addition, other functions can be added on top to fit models with covariance structures, i.e. the `g()` function which indicates that the random effect inside the parenthesis (i.e. GCA2) has a covariance matrix (A, pedigree or genomic relationship matrix) that will be specified in the G argument in the form of a list:

```{r}
data(cornHybrid)
hybrid2 <- cornHybrid$hybrid # extract cross data
## get the covariance structure for GCA2
A <- cornHybrid$K
## fit the model
modFD <- mmer2(Yield~1, 
               random=~ g(GCA2) + at(Location):g(GCA2), 
               rcov= ~ at(Location):units,
               data=hybrid2, G=list(GCA2=A),
               silent = TRUE, draw=FALSE)
summary(modFD)
```

The `draw` argument allows you to see the progress of the likelihood and the change of the variance components, we just mention it in case you like to do that inspection but this will make the fitting process more time consuming.

## 3) Using the pin calculator

Sometimes the user needs to calculate ratios or functions of specific variance-covariance components and obtain the standard error for such parameters. Examples of these are the genetic correlations, heritabilities, etc. Using the CPdata we will show how to estimate the heritability and the standard error.

```{r}
data(CPdata)
#### create the variance-covariance matrix 
A <- A.mat(CPgeno)
#### look at the data and fit the model
head(CPpheno)
mix1 <- mmer2(color~1,
              random=~g(id), 
              rcov=~units,
              G=list(id=A), data=CPpheno, silent=TRUE)
summary(mix1)
#### run the pin function
pin(mix1, h2 ~ V1 / ( V1 + V2 ) )
```

The same can be used for multivariate models. Please check the documentation of the `pin` function to see more examples.

## 4) Half and full diallel designs 

When breeders are looking for the best single cross combinations, diallel designs have been by far the most used design in crops like maize. There are 4 types of diallel designs depending if reciprocate and self cross (omission of parents) are performed (full diallel with parents n^2; full diallel without parents n(n-1); half diallel with parents 1/2 * n(n+1); half diallel without parents 1/2 * n(n-1) ). In this example we will show a full dialle design (reciprocate crosses are performed) and half diallel designs (only one of the directions is performed).

In the first data set we show a full diallel among 40 lines from 2 heterotic groups, 20 in each. Therefore 400 possible hybrids are possible. We have pehnotypic data for 100 of them across 4 locations. We use the data available to fit a model of the form:

<br>

$y = X\beta + Zu_1 + Zu_2 + Zu_S + \epsilon$ 

<br>

We estimate variance components for $GCA_1$, $GCA_2$ and $SCA$ and use them to estimate heritability. Additionally BLUPs for GCA and SCA effects can be used to predict crosses.

```{r}
data(cornHybrid)
hybrid2 <- cornHybrid$hybrid # extract cross data
head(hybrid2)

modFD <- mmer2(Yield~Location, 
               random=~GCA1+GCA2+SCA, 
               rcov=~units,
               data=hybrid2,silent = TRUE, draw=FALSE)
(suma <- summary(modFD))
Vgca <- sum(suma$var.comp.table[1:2,1])
Vsca <- suma$var.comp.table[3,1]
Ve <- suma$var.comp.table[4,1]
Va = 4*Vgca
Vd = 4*Vsca
Vg <- Va + Vd
(H2 <- Vg / (Vg + (Ve)) )
(h2 <- Va / (Vg + (Ve)) )
```
Don't worry too much about the small h2 value, the data was simulated to be mainly dominance variance, therefore the Va was simulated extremely small leading to such value of narrow sense h2.

In this second data set we show a small half diallel with 7 parents crossed in one direction. n(n-1)/2 crosses are possible 7(6)/2 = 21 unique crosses. Parents appear as males or females indistictly. Each with two replications in a CRD. For a half diallel design a single GCA variance component for both males and females can be estimated and an SCA as well ($\sigma^2_GCA$ and $\sigma^2_SCA$ respectively), and BLUPs for GCA and SCA of the parents can be extracted. We would show first how to use it with the `mmer2` function using the `and()` function and later we will show how to do it creating customized matrices using the `overlay` and `model.matrix` functions for the GCA and SCA matrices respectively. The specific model here is:

$y = X\beta + Zu_g + Zu_s + \epsilon$ 

```{r}
data(HDdata)
head(HDdata)
HDdata$geno <- as.factor(HDdata$geno)
HDdata$male <- as.factor(HDdata$male)
HDdata$female <- as.factor(HDdata$female)
# Fit the model
modHD <- mmer2(sugar~1, 
               random=~male + and(female) + geno, 
               rcov=~units,
               data=HDdata, silent = TRUE)
summary(modHD)
suma <- summary(modHD)$var.comp.table
Vgca <- suma[1,1]
Vsca <- suma[2,1]
Ve <- suma[3,1]
Va = 4*Vgca
Vd = 4*Vsca
Vg <- Va + Vd
(H2 <- Vg / (Vg + (Ve/2)) ) # 2 technical reps
(h2 <- Va / (Vg + (Ve/2)) )
```

Notice how the `and()` argument makes the overlay possible making sure that male and female are joint into a single random effect. The same can be done using the `mmer` argument by creating the incidence and covariance matrices in case you want to see what is doing `mmer2` in the background.

```{r}
  data(HDdata)
  head(HDdata)
  #### GCA matrix for half diallel using male and female columns
  #### use the 'overlay' function to create the half diallel matrix
  Z1 <- overlay(HDdata[,c(3:4)])
  #### Obtain the SCA matrix
  Z2 <- model.matrix(~as.factor(geno)-1, data=HDdata)
  #### Define the response variable and run
  y <- HDdata$sugar
  ETA <- list(list(Z=Z1), list(Z=Z2)) # Zu component
  modHD <- mmer(Y=y, Z=ETA, draw=FALSE, silent=TRUE)
  summary(modHD)
```

## 5) Genomic selection

In this section we will use wheat data from CIMMYT to show how is genomic selection performed. This is the case of prediction of specific individuals within a population. It basically uses a similar model of the form:

<br>

$y = X\beta + Zu + \epsilon$ 

<br>

and takes advantage of the variance covariance matrix for the genotype effect known as the additive relationship matrix (A) and calculated using the `A.mat` function to establish connections among all individuals and predict the BLUPs for individuals that were not measured. The prediction accuracy depends on several factors such as the heritability ($h^2$), training population used (TP), size of TP, etc.

```{r}
data(wheatLines); 
X <- wheatLines$wheatGeno; X[1:5,1:4]; dim(X)
Y <- data.frame(wheatLines$wheatPheno); Y$id <- rownames(Y); head(Y);
rownames(X) <- rownames(Y)
# select environment 1
K <- A.mat(X) # additive relationship matrix
# GBLUP pedigree-based approach
set.seed(12345)
y.trn <- Y
vv <- sample(rownames(Y),round(dim(Y)[1]/5))
y.trn[vv,"X1"] <- NA
ans <- mmer2(X1~1,
             random=~g(id), 
             rcov=~units,
             G=list(id=K), method="NR", 
             data=y.trn, silent = TRUE) # kinship based
cor(ans$u.hat$`g(id)`[vv,],Y[vv,"X1"])
```

## 6) Single cross prediction

When doing prediction of single cross performance the phenotype can be dissected in three main components, the general combining abilities (GCA) and specific combining abilities (SCA). This can be expressed with the same model analyzed in the diallel experiment mentioned before:

<br>

$y = X\beta + Zu_1 + Zu_2 + Zu_S + \epsilon$ 

<br>

with:

<br>

$u_1$ ~ N(0, $K_1$$\sigma^2_u1$)

$u_2$ ~ N(0, $K_2$$\sigma^2_u2$)

$u_s$ ~ N(0, $K_3$$\sigma^2_us$)

<br>

And we can specify the K matrices. The main difference between this model and the full and half diallel designs is the fact that this model will include variance covariance structures in each of the three random effects (GCA1, GCA2 and SCA) to be able to predict the crosses that have not ocurred yet. We will use the data published by Technow et al. (2015) to show how to do prediction of single crosses.

```{r}
data(Technow_data)

A.flint <- Technow_data$AF # Additive relationship matrix Flint
A.dent <- Technow_data$AD # Additive relationship matrix Dent

pheno <- Technow_data$pheno # phenotypes for 1254 single cross hybrids
head(pheno);dim(pheno) 
# CREATE A DATA FRAME WITH ALL POSSIBLE HYBRIDS
DD <- kronecker(A.dent,A.flint,make.dimnames=TRUE)
hybs <- data.frame(sca=rownames(DD),yield=NA,matter=NA,gcad=NA, gcaf=NA)
hybs$yield[match(pheno$hy, hybs$sca)] <- pheno$GY
hybs$matter[match(pheno$hy, hybs$sca)] <- pheno$GM
hybs$gcad <- as.factor(gsub(":.*","",hybs$sca))
hybs$gcaf <- as.factor(gsub(".*:","",hybs$sca))
head(hybs)
# RUN THE PREDICTION MODEL
y.trn <- hybs
vv1 <- which(!is.na(hybs$yield))
vv2 <- sample(vv1, 100)
y.trn[vv2,"yield"] <- NA
anss2 <- mmer2(yield~1, 
               random=~g(gcad) + g(gcaf), 
               rcov=~units,
               G=list(gcad=A.dent, gcaf=A.flint), 
               method="NR", silent=TRUE, data=y.trn) 
summary(anss2)
cor(anss2$fitted.y[vv2], hybs$yield[vv2])
```

In the previous model we only used the GCA effects (GCA1 and GCA2) for practicity, altough it's been shown that the SCA effect doesn't actually help that much in increasing prediction accuracy and increase a lot the computation intensity required since the variance covariance matrix for SCA is the kronecker product of the variance covariance matrices for the GCA effects, resulting in a 10578x10578 matrix that increases in a very intensive manner the computation required.

A model without covariance structures would show that the SCA variance component is insignificant compared to the GCA effects. This is why including the third random effect doesn't increase the prediction accuracy.

## 7) Multivariate genetic models and genetic correlations

Sometimes is important to estimate genetic variance-covariance among traits, multi-reponse models are very useful for such task. Let see an example with 3 traits (color, Yield, and Firmness) and a single random effect (genotype; id) although multiple effects can be modeled as well. We need to use a variance covariance structure for the random effect to be able to obtain the genetic covariance among traits.

```{r}
data(CPdata)
### look at the data
head(CPpheno);CPgeno[1:5,1:4]
## fit a model including additive effects
A <- A.mat(CPgeno) # additive relationship matrix
####================####
#### ADDITIVE MODEL ####
####================####
ans.A <- mmer2(cbind(color,Yield)~1, 
               random=~us(trait):g(id),
               rcov=~us(trait):units,
               G=list(id=A),
               data=CPpheno, silent = TRUE)
summary(ans.A)
```

Now you can extract the BLUPs using the 'randef' function or simple accesing with the '$' sign and pick 'u.hat'. Also, genetic correlations and heritabilities can be calculated easily. 

```{r}
## genetic variance covariance
gvc <- ans.A$var.comp$`g(id)`
## extract variances (diagonals) and get standard deviations
sd.gvc <- as.matrix(sqrt(diag(gvc))) 
## get possible products sd(Vgi) * sd(Vgi')
prod.sd <- sd.gvc %*% t(sd.gvc)
## genetic correlations cov(gi,gi')/[sd(Vgi) * sd(Vgi')]
(gen.cor <- gvc/prod.sd)
## heritabilities
(h2 <- diag(gvc) / diag(cov(CPpheno[,names(diag(gvc))], use = "complete.obs")))
```

## Literature

Covarrubias-Pazaran G. 2016. Genome assisted prediction of quantitative traits using the R package sommer. PLoS ONE 11(6):1-15.

Bernardo Rex. 2010. Breeding for quantitative traits in plants. Second edition. Stemma Press. 390 pp.

Gilmour et al. 1995. Average Information REML: An efficient algorithm for variance parameter estimation in linear mixed models. Biometrics 51(4):1440-1450.

Henderson C.R. 1975. Best Linear Unbiased Estimation and Prediction under a Selection Model. Biometrics vol. 31(2):423-447.

Kang et al. 2008. Efficient control of population structure in model organism association mapping. Genetics 178:1709-1723.

Lee et al. 2015. MTG2: An efficient algorithm for multivariate linear mixed model analysis based on genomic information. Cold Spring Harbor. doi: http://dx.doi.org/10.1101/027201.

Searle. 1993. Applying the EM algorithm to calculating ML and REML estimates of variance components. Paper invited for the 1993 American Statistical Association Meeting, San Francisco.

Yu et al. 2006. A unified mixed-model method for association mapping that accounts for multiple levels of relatedness. Genetics 38:203-208.

Abdollahi Arpanahi R, Morota G, Valente BD, Kranis A, Rosa GJM, Gianola D. 2015. Assessment of bagging GBLUP for whole genome prediction of broiler chicken traits. Journal of Animal Breeding and Genetics 132:218-228.

Tunnicliffe W. 1989. On the use of marginal likelihood in time series model estimation. JRSS 51(1):15-27.